Finally Tagless, Partially Evaluated

### 2007-07-27 tagless/tagless-final.tex
small edits here and there

We can modify the CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state and implement an extended object language, with operations
|deref ()| to obtain the current state, |set v| to replace the current
state with |v| returning the previous value of the state, and the
let-form |let x = e1 in e2|. The latter enforces the order of
evaluation: |e1| will always be evaluated before |e2| even if the
expression |e2| does not need the value of |x|. The more familiar
imperative sequencing form |begin e1; e2 end| is expressible in our
object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
has the form |lapp e1 (fun x -> e2)|, which looks like the functional
application with its two arguments swapped.  The extended object
language corresponds to the following signature, which
includes all of the original Symantics and adds imperative features.

\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states	(* static version of the state *)
  val lapp : ('c,'sa,'da) repr -> 
             (('c,'sa,'da) repr -> ('c,'sb,'db) repr) ->
             ('c,'sb,'db) repr
  val deref : unit -> ('c,'c states,state) repr
  val set   : (('c,'c states,state) repr as 'x) -> 'x
end
\end{code}

We can write a simple test: 
\begin{code}
let v0 = !state in state := 2; v0 + !state
\end{code}
which, in our language, takes the form
\begin{code}
module EXSI_INT(S: SymSI with type state = int and type 'c states = int) = 
struct
  open S
  let test1 () = lapp (deref ()) (fun v0 -> 
                  lapp (set (int 2)) (fun _ ->
		   add v0 (deref ())))
end;;
\end{code}
The accompanying source code shows several more tests, including the
imperative power function that uses the state as the accumulator and
the test of a higher-order state.

The state-passing interpreter is the extension of the CBN CPS
interpreter RCN of\S\ref{S:CPS}:
\begin{code}
module RCPS (ST: sig 
  type state 
  type 'c states 
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end) = struct
  include ST
  type ('c, 'sv, 'dv) result = 'c states -> 'sv
  ... 
  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
  let lapp e2 e1 = 
    {ko = fun k -> e2.ko (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
  let deref () = {ko = fun k s -> k s s}
  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
end
\end{code}
The implementations of |int|, |app|, |lam|, etc. functions is \emph{identical}
to those of |RCN| and elided. New here is the extended type |repr|
that now includes the state, and the `constructors' of the imperative
forms of the object language: |lapp|, |deref|, and |set|. The
interpreter is still CBN and so evaluation of |app ef ea| does not
guarantee to evaluate |ea|. On the other hand, the implementation of
the let-form |lapp ea ef| does assure that |ea| will always be
evaluated. In the case of the first-order state such as |int| we 
instantiate the interpreter as
\begin{code}
module RCPSI = RCPS(struct 
  type state = int
  type 'c states = int
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end);;
\end{code}
It fully implements the |Symantics| signature and so can run the pure
code of the unextended object language: we can pass |RCPSI| as an
argument to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to run the 
extended object language:
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100 (* 102 *)
\end{code}

We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.  No new techniques were needed in the implementation.

@@

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \cite{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.

@@

It is common to implement an embedded DSL by providing multiple
interpretations of host\hyp language pervasives such as addition and
application.  It is also common to use phantom types to rule out
ill-typed object terms, as Lava
\cite{Lava} and \citet{Rhiger-thesis} do. However, these approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig} and Rhiger's \texttt{Raw} (his Fig.~2.2)
and \texttt{Term} (his Chap.~3), and incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\cite{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}

@@

We encode terms in elimination form, as a coalgebraic structure.
\Citet{Pfenning-Lee} first described this basic idea and applied it to
metacircular interpretation.
Our approach, however, can be implemented in mainstream ML and supports
type inference and typed CPS transformation and partial evaluation. In contrast,
\citeauthor{Pfenning-Lee} conclude that partial evaluation and program
transformations ``do not seem to be expressible'' even using their
extension to~$F_\omega$, perhaps because their avoidance of general
recursive types compels them to include the polymorphic lift that we
avoid in~\S\ref{S:PE-lift}.
\begin{comment}
It seems that Pfenning and Lee embed $F_2$ with type constructions in 
(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
we do is very similar to what they do (Figure 1, p.152), except that we 
do it in standard programming languages.  It is unclear if their work can 
be implemented (yet) in any language.  And we preserve type-inference, 
while their solution needs explicit types!
The following line of their conclusion is worth citing: "... this does 
not imply that the same language is also suitable for type 
metaprogramming. ... such as partial evaluation... do not seem to be 
expressible".
I suspect you're right, but I'm still reading the paper.  See also page
146: "for a term M in $F_1$ (a simply-typed term), the representation
$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
and beyond reminds me
strongly of our attempts at self-interpretation without the notion of a
syntactic hole.
\end{comment}

@@

\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
higher-rank and higher-kind
polymorphism let us type-check and compile object terms separately from
interpreters.  This \ifshort\else observation \fi is consistent with the role of
polymorphism in the separate compilation of modules
\cite{shao-typed}.

\begin{comment}
Our solution could have been even more straightforward if we had
total type-level functions at our disposal (|repr_pe| in
\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
certainly ought to become mainstream.
\end{comment}

>>>>

We can modify the CBV CPS transformation to pass a piece of state
along with the continuation. This technique lets us support mutable
state. This extended object language adds operations
|deref ()| to obtain the current state, |set v| to replace the current
state with |v| returning the previous value of the state, and the
let-form |let x = e1 in e2|. The latter enforces an order of
evaluation: |e1| will always be evaluated before |e2| even if the
expression |e2| does not need the value of |x|. The more familiar
imperative sequencing form |begin e1; e2 end| is expressible in our
object language as |let _ = e1 in e2|.  In HOAS, the let-form actually
has the form |lapp e1 (fun x -> e2)|, which looks like functional
application with its two arguments swapped.  The extended object
language corresponds to the following signature, which
includes all of the original |Symantics| and adds imperative features.

\begin{code}
module type SymSI = sig
  include Symantics
  type state
  type 'c states       (* static version of the state *)
  val lapp : ('c,'sa,'da) repr -> 
             (('c,'sa,'da) repr -> ('c,'sb,'db) repr) ->
             ('c,'sb,'db) repr
  val deref : unit -> ('c,'c states,state) repr
  val set   : (('c,'c states,state) repr as 'x) -> 'x
end
\end{code}
We can write a simple test: 
\begin{code}
let v0 = !state in state := 2; v0 + !state
\end{code}
which, in our language, takes the form
\begin{code}
module EXSI_INT(S: SymSI with type state = int 
                          and type 'c states = int) = 
struct
  open S
  let test1 () = lapp (deref ()) (fun v0 -> 
                  lapp (set (int 2)) (fun _ ->
                   add v0 (deref ())))
end;;
\end{code}
The accompanying source code shows several more tests, including an
imperative power function that uses state as the accumulator as well as
a test for a higher-order state.

The state-passing interpreter is an extension of the CBN CPS
interpreter RCN of\S\ref{S:CPS}:
\begin{code}
module RCPS (ST: sig 
  type state 
  type 'c states 
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end) = struct
  include ST
  type ('c, 'sv, 'dv) result = 'c states -> 'sv
  ... 
  let get_res x = fun s0 -> x.ko (fun v s -> v) s0
  let lapp e2 e1 = {ko = fun k -> e2.ko 
      (fun v -> (app (lam e1) {ko = fun k -> k v}).ko k)}
  let deref () = {ko = fun k s -> k s s}
  let set e = {ko = fun k -> e.ko (fun v s -> k s v)}
end
\end{code}
The implementations of the |int|, |app|, |lam|, etc. functions is
\emph{identical} to those of |RCN| and elided. New here is the extended type
|repr|
which now includes the state, and the ``constructors'' for the imperative
forms of the object language: |lapp|, |deref|, and |set|. The
interpreter is still CBN and so evaluation of |app ef ea| does not
guarantee to evaluate |ea|. On the other hand, the implementation of
the let-form |lapp ea ef| does assure that |ea| will always be
evaluated. In the case of first-order state (such as |int|) we 
instantiate the interpreter as
\begin{code}
module RCPSI = RCPS(struct 
  type state = int
  type 'c states = int
  type ('c,'sv,'dv) repr = 
      {ko: 'w. ('sv -> 'c states -> 'w) -> 'c states -> 'w}
end);;
\end{code}
It fully implements the |Symantics| signature and so can run the pure
code of the unextended object language: we can pass |RCPSI| as an
argument to the functor |EX| (Fig.~\ref{fig:ocaml-simple}) and run 
the example |test1| from there. The main use for |RCPSI| is to run the 
extended object language:
\begin{code}
module EXPSI_INT = EXSI_INT(RCPSI)
let cpsitesti1 = RCPSI.get_res (EXPSI_INT.test1 ()) 100
val cpsitesti1 : int = 102
\end{code}

We can also add mutable references to the object language using
mutable references of the metalanguage, as shown in the accompanying
code.  It is worthwhile reiterating how close this implementation is to the CPS
interpreter, and that adding state and imperative features needed no new
techniques.

@@

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \cite{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
\begin{comment}
Our solution could have been even more straightforward if we had
total type-level functions at our disposal (|repr_pe| in
\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
certainly ought to become mainstream.
\end{comment}

@@

It is common to implement an embedded DSL by providing multiple
interpretations of host\hyp language pervasives such as addition and
application.  It is also common to use phantom types to rule out
ill-typed object terms, as done in Lava
\cite{Lava} and by \citet{Rhiger-thesis}. However, these approaches
are not tagless because they still use universal types, such as Lava's
\texttt{Bit} and \texttt{NumSig}, and Rhiger's \texttt{Raw} (his Fig.~2.2)
and \texttt{Term} (his Chap.~3), which incur the attendant overhead of
pattern matching.  The universal type also
greatly complicates the soundness and completeness proofs of embedding
\cite{Rhiger-thesis}, whereas our proofs are trivial.
Rhiger's approach does not support typed CPS transformation (his~\S3.3.4).
\begin{comment}
Rhiger's But Fig 2.2, p33: universal type Raw.  He uses phantom type
upon the Exp datatype. But that is cheating: phantom type means
essentially we can easily do coerce. We use real types.  That's why he
had to do tedious proofs in Sec 2 of soundness and completeness of
embedding. Whereas our proofs are obvious.  His sec 3 is based on data
representation of terms. They have type tags.  We do nothing of that
kind: See Sec 3.1.2. See numerous "data Term" in Sec3, which is the U
type.  In Sec 3.3.4 (p76) Rhiger specifically says that his encoding
cannot do typed CPS transformation -- whereas our does. BTW, Rhiger
thesis contains the definitions of the interpreter and the compiler,
in the beginning. Use this in response to Rev1)
\end{comment}

@@

We encode terms in elimination form, as a coalgebraic structure.
\Citet{Pfenning-Lee} first described this basic idea and applied it to
metacircular interpretation.
Our approach, however, can be implemented in mainstream ML and supports
type inference, typed CPS transformation and partial evaluation. In contrast,
\citeauthor{Pfenning-Lee} conclude that partial evaluation and program
transformations ``do not seem to be expressible'' even using their
extension to~$F_\omega$, perhaps because their avoidance of general
recursive types compels them to include the polymorphic lift that we
avoid in~\S\ref{S:PE-lift}.
\begin{comment}
It seems that Pfenning and Lee embed $F_2$ with type constructions in 
(pure) $F_3$.  We embed $F_1$ in (weak?) $F_2$, as I see it.  In a way, what 
we do is very similar to what they do (Figure 1, p.152), except that we 
do it in standard programming languages.  It is unclear if their work can 
be implemented (yet) in any language.  And we preserve type-inference, 
while their solution needs explicit types!
The following line of their conclusion is worth citing: "... this does 
not imply that the same language is also suitable for type 
metaprogramming. ... such as partial evaluation... do not seem to be 
expressible".
I suspect you're right, but I'm still reading the paper.  See also page
146: "for a term M in $F_1$ (a simply-typed term), the representation
$\bar{M}$ will be in $F_2$".  The move from $F_2$ to $F_3$ 
and beyond reminds me
strongly of our attempts at self-interpretation without the notion of a
syntactic hole.
\end{comment}

@@

\ifshort We observe that \else As \S\ref{S:clarify} observes, \fi
higher-rank and higher-kind
polymorphism lets us type-check and compile object terms separately from
interpreters.  This \ifshort\else observation \fi is consistent with the role of
polymorphism in the separate compilation of modules
\cite{shao-typed}.
